### 2018-07-11 submission/manuscript.Rmd
intro edits

The bacterial community is reported to vary between case and control for a number of diseases [@Turnbaugh2008; @Baxter2016; @Sze2015; @Bonfili2017]. However, for diseases like obesity, the taxa identified have varied widely depending on the study [@Turnbaugh2008; @Zupancic2012]. Some of this variation could be due to error introduced during the 16S rRNA gene sequencing workflow. A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps. The preservation and extraction stages of the 16S rRNA gene sequencing workflow have been the most extensively studied [@contamination_Salter2014; @preservation_Song_2016; @storage_Bassis_2017; @review_Kim_2017]. For preservation and extraction stages of the workflow, it has been consistently found that there are errors based on the kits used, but that these differences are smaller than the overall biological difference measured between samples with different kits [@preservation_Song_2016; @storage_Bassis_2017]. Since these studies use the same PCR approach while varying preservation or extraction method, the contribution of PCR error to this overall workflow is not well characterized.

There is a large body of literature that shows there are errors due to primer and number of cycles chosen for the PCR stage of 16S rRNA gene sequencing [@Eckert1991; @Burkardt2000]. Primers have variable region dependent binding affinities which causes an inability to detect specific bacteria (e.g. V1-V3 does not detect *Haemophilus influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [@Meisel2016; @Sze2015 (Table S4)]. Another source of error is the selective amplification of AT-rich over GC-rich sequences which exaggerate the difference between 16S rRNA genes higher in AT versus those higher in GC content [@polz_bias_1998]. Many of these sources of error are made worse as the number of cycles increases [@Wang1996; @Haas2011; @Kebschull2015]. For example, both amplification error and non-specific amplification (e.g. incorrect amplicon size products) also can increase as the number of cycles increases. This will increase the number of Operational Taxonomic Units (OTUs) observed and drastically change the values obtained from commonly used diversity measures [@BautistadelosSantos2016; @Acinas2005]. Additionally, as the number of cycles increases more chimeras can form from an aborted extension step that causes a priming error and subsequent secondary extension [@Haas2011]. These chimeras will artificially increase community diversity by increasing the number of OTUs that are observed [@Haas2011]. In addition to these sources of errors, there also are multiple families of DNA polymerases that have their own error rate and proof reading capacity [@polymerase_Ishino_2014]. Interestingly, the influence that these different DNA polymerases can have on the observed 16S rRNA gene sequencing results have not been well studied like some of the other sources of PCR-based bias.

A recent study found differences in the number of OTUS and chimeras between normal and high fidelity DNA polymerases [@taq_Gohl_2016]. The authors could reduce the difference between the two polymerases by optimizing the annealing and extension steps within the PCR protocol [@taq_Gohl_2016]. Yet, within this study there was no comparison made between different high fidelity DNA polymerases. Additionally, the more complex fecal samples that were analyzed had different methodolgies applied to the two different groups tested, which could exagerate the actual biological differences detected. Due to this gap, it is natural to extend this line of inquiry. First, for detecting whether there are high fidelity polymerase and number of cycle dependent error. Second, for examining whether these differences are enough to obscure smaller biological signals than what has been previously reported [@taq_Gohl_2016]. Thus, this study will investigate whether differences in high fidelity DNA polymerases are due to error and whether these differences obscure the biological variation in the fecal bacterial community composition between individuals. We will accomplish this by examining GC-based amplification error, sequence error (e.g. subsitutions), chimera prevalence, the number of OTUs, Bray-Curtis index, and OTU-based differences at varying number of cycles in five different high fidelity DNA polymerases. Additionally, we also used Random Forest models to classify number of cycles, polymerase, and individuals to assess which model was best when using 16S rRNA gene sequencing data.

>>>>

The bacterial community is reported to vary between case and control for a number of diseases [@Turnbaugh2008; @Baxter2016; @Sze2015; @Bonfili2017]. However, for diseases like obesity, the taxa identified have varied widely depending on the study [@Turnbaugh2008; @Zupancic2012]. Some of this variation could be due to error introduced during the 16S rRNA gene sequencing workflow. A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps. The preservation and extraction stages of the 16S rRNA gene sequencing workflow have been the most extensively studied [@contamination_Salter2014; @preservation_Song_2016; @storage_Bassis_2017; @review_Kim_2017]. For preservation and extraction stages of the workflow, it has been consistently found that there are errors in the observed bacterial community composition based on the kits used, but that these differences are smaller than the overall biological difference measured between samples with different kits [@preservation_Song_2016; @storage_Bassis_2017]. Since these studies use the same PCR approach while varying preservation or extraction method, the contribution of PCR error to this overall workflow is not well characterized.

There is a large body of literature that shows there are errors due to primer and number of cycles chosen for the PCR stage of 16S rRNA gene sequencing [@Eckert1991; @Burkardt2000]. Primers have variable region dependent binding affinities which causes an inability to detect specific bacteria (e.g. V1-V3 does not detect *Haemophilus influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [@Meisel2016; @Sze2015 (Table S4)]. Another source of error is the selective amplification of AT-rich over GC-rich sequences which exaggerate the difference between 16S rRNA genes higher in AT versus those higher in GC content [@polz_bias_1998]. Many of these sources of error are made worse as the number of cycles increases [@Wang1996; @Haas2011; @Kebschull2015]. For example, both amplification error and non-specific amplification (e.g. incorrect amplicon size products) also can increase as the number of cycles increases. This will increase the number of Operational Taxonomic Units (OTUs) observed and drastically change the values obtained from commonly used diversity measures [@BautistadelosSantos2016; @Acinas2005]. Additionally, as the number of cycles increases more chimeras can form from an aborted extension step that causes a priming error and subsequent secondary extension [@Haas2011]. These chimeras will artificially increase community diversity by increasing the number of OTUs that are observed [@Haas2011]. In addition to these sources of errors, there also are multiple families of DNA polymerases that have their own error rate and proof reading capacity [@polymerase_Ishino_2014]. Interestingly, the influence that these different DNA polymerases can have on the observed 16S rRNA gene sequencing results have not been well studied like some of the other sources of PCR-based error

A recent study found differences in the number of OTUS and chimeras between normal and high fidelity DNA polymerases [@taq_Gohl_2016]. The authors of this study could reduce the difference between two polymerases by optimizing the annealing and extension steps within the PCR protocol [@taq_Gohl_2016]. Yet, within this study there was no comparison made between different high fidelity DNA polymerases on the same samples. Additionally, the fecal samples that were analyzed, as a model of a more complex community, had different processing methodolgies applied to the two different treatment groups tested, which could exagerate the observed biological differences. Due to these gaps, it is natural to extend the previous line of inquiry to include a more detailed examination of two specific areas. First, for the investigation of whether differences in PCR error are detected between high fidelity polymerases and number of cycles used. Second, for examining whether these differences in error are enough to obscure smaller biological signals than what has been previously reported [@taq_Gohl_2016]. Thus, this study will investigate whether differences in high fidelity DNA polymerases are due to error and whether these differences obscure the biological variation in the fecal bacterial community composition between individuals. We accomplish this by examining GC-based amplification error, sequence error (e.g. subsitutions), chimera prevalence, the number of OTUs, Bray-Curtis index, and OTU-based differences at varying number of cycles in five different high fidelity DNA polymerases. Additionally, we also created Random Forest models to classify number of cycles, polymerase, and individuals to assess which model had the best classification success when using 16S rRNA gene sequencing data.

### 2018-07-11 submission/manuscript.tex
intro edits

The bacterial community is reported to vary between case and control for
a number of diseases (Turnbaugh et al., 2008; Sze et al., 2015; Baxter
et al., 2016; Bonfili et al., 2017). However, for diseases like obesity,
the taxa identified have varied widely depending on the study (Turnbaugh
et al., 2008; Zupancic et al., 2012). Some of this variation could be
due to error introduced during the 16S rRNA gene sequencing workflow. A
typical 16S rRNA gene sequencing workflow can be divided into
preservation, extraction, PCR, and sequencing steps. The preservation
and extraction stages of the 16S rRNA gene sequencing workflow have been
the most extensively studied (Salter et al., 2014; Song et al., 2016;
Bassis et al., 2017; Kim et al., 2017). For preservation and extraction
stages of the workflow, it has been consistently found that there are
errors based on the kits used, but that these differences are smaller
than the overall biological difference measured between samples with
different kits (Song et al., 2016; Bassis et al., 2017). Since these
studies use the same PCR approach while varying preservation or
extraction method, the contribution of PCR error to this overall
workflow is not well characterized.

There is a large body of literature that shows there are errors due to
primer and number of cycles chosen for the PCR stage of 16S rRNA gene
sequencing (Eckert \& Kunkel, 1991; Burkardt, 2000). Primers have
variable region dependent binding affinities which causes an inability
to detect specific bacteria (e.g.~V1-V3 does not detect
\emph{Haemophilus influenzae} and V3-V5 does not detect
\emph{Propionibacterium acnes}) (Sze et al., 2015 (Table S4); Meisel et
al., 2016). Another source of error is the selective amplification of
AT-rich over GC-rich sequences which exaggerate the difference between
16S rRNA genes higher in AT versus those higher in GC content (Polz \&
Cavanaugh, 1998). Many of these sources of error are made worse as the
number of cycles increases (Wang \& Wang, 1996; Haas et al., 2011;
Kebschull \& Zador, 2015). For example, both amplification error and
non-specific amplification (e.g.~incorrect amplicon size products) also
can increase as the number of cycles increases. This will increase the
number of Operational Taxonomic Units (OTUs) observed and drastically
change the values obtained from commonly used diversity measures (Acinas
et al., 2005; Santos et al., 2016). Additionally, as the number of
cycles increases more chimeras can form from an aborted extension step
that causes a priming error and subsequent secondary extension (Haas et
al., 2011). These chimeras will artificially increase community
diversity by increasing the number of OTUs that are observed (Haas et
al., 2011). In addition to these sources of errors, there also are
multiple families of DNA polymerases that have their own error rate and
proof reading capacity (Ishino \& Ishino, 2014). Interestingly, the
influence that these different DNA polymerases can have on the observed
16S rRNA gene sequencing results have not been well studied like some of
the other sources of PCR-based bias.

A recent study found differences in the number of OTUS and chimeras
between normal and high fidelity DNA polymerases (Gohl et al., 2016).
The authors could reduce the difference between the two polymerases by
optimizing the annealing and extension steps within the PCR protocol
(Gohl et al., 2016). Yet, within this study there was no comparison made
between different high fidelity DNA polymerases. Additionally, the more
complex fecal samples that were analyzed had different methodolgies
applied to the two different groups tested, which could exagerate the
actual biological differences detected. Due to this gap, it is natural
to extend this line of inquiry. First, for detecting whether there are
high fidelity polymerase and number of cycle dependent error. Second,
for examining whether these differences are enough to obscure smaller
biological signals than what has been previously reported (Gohl et al.,
2016). Thus, this study will investigate whether differences in high
fidelity DNA polymerases are due to error and whether these differences
obscure the biological variation in the fecal bacterial community
composition between individuals. We will accomplish this by examining
GC-based amplification error, sequence error (e.g.~subsitutions),
chimera prevalence, the number of OTUs, Bray-Curtis index, and OTU-based
differences at varying number of cycles in five different high fidelity
DNA polymerases. Additionally, we also used Random Forest models to
classify number of cycles, polymerase, and individuals to assess which
model was best when using 16S rRNA gene sequencing data.

>>>>

The bacterial community is reported to vary between case and control for
a number of diseases (Turnbaugh et al., 2008; Sze et al., 2015; Baxter
et al., 2016; Bonfili et al., 2017). However, for diseases like obesity,
the taxa identified have varied widely depending on the study (Turnbaugh
et al., 2008; Zupancic et al., 2012). Some of this variation could be
due to error introduced during the 16S rRNA gene sequencing workflow. A
typical 16S rRNA gene sequencing workflow can be divided into
preservation, extraction, PCR, and sequencing steps. The preservation
and extraction stages of the 16S rRNA gene sequencing workflow have been
the most extensively studied (Salter et al., 2014; Song et al., 2016;
Bassis et al., 2017; Kim et al., 2017). For preservation and extraction
stages of the workflow, it has been consistently found that there are
errors in the observed bacterial community composition based on the kits
used, but that these differences are smaller than the overall biological
difference measured between samples with different kits (Song et al.,
2016; Bassis et al., 2017). Since these studies use the same PCR
approach while varying preservation or extraction method, the
contribution of PCR error to this overall workflow is not well
characterized.

There is a large body of literature that shows there are errors due to
primer and number of cycles chosen for the PCR stage of 16S rRNA gene
sequencing (Eckert \& Kunkel, 1991; Burkardt, 2000). Primers have
variable region dependent binding affinities which causes an inability
to detect specific bacteria (e.g.~V1-V3 does not detect
\emph{Haemophilus influenzae} and V3-V5 does not detect
\emph{Propionibacterium acnes}) (Sze et al., 2015 (Table S4); Meisel et
al., 2016). Another source of error is the selective amplification of
AT-rich over GC-rich sequences which exaggerate the difference between
16S rRNA genes higher in AT versus those higher in GC content (Polz \&
Cavanaugh, 1998). Many of these sources of error are made worse as the
number of cycles increases (Wang \& Wang, 1996; Haas et al., 2011;
Kebschull \& Zador, 2015). For example, both amplification error and
non-specific amplification (e.g.~incorrect amplicon size products) also
can increase as the number of cycles increases. This will increase the
number of Operational Taxonomic Units (OTUs) observed and drastically
change the values obtained from commonly used diversity measures (Acinas
et al., 2005; Santos et al., 2016). Additionally, as the number of
cycles increases more chimeras can form from an aborted extension step
that causes a priming error and subsequent secondary extension (Haas et
al., 2011). These chimeras will artificially increase community
diversity by increasing the number of OTUs that are observed (Haas et
al., 2011). In addition to these sources of errors, there also are
multiple families of DNA polymerases that have their own error rate and
proof reading capacity (Ishino \& Ishino, 2014). Interestingly, the
influence that these different DNA polymerases can have on the observed
16S rRNA gene sequencing results have not been well studied like some of
the other sources of PCR-based error

A recent study found differences in the number of OTUS and chimeras
between normal and high fidelity DNA polymerases (Gohl et al., 2016).
The authors of this study could reduce the difference between two
polymerases by optimizing the annealing and extension steps within the
PCR protocol (Gohl et al., 2016). Yet, within this study there was no
comparison made between different high fidelity DNA polymerases on the
same samples. Additionally, the fecal samples that were analyzed, as a
model of a more complex community, had different processing methodolgies
applied to the two different treatment groups tested, which could
exagerate the observed biological differences. Due to these gaps, it is
natural to extend the previous line of inquiry to include a more
detailed examination of two specific areas. First, for the investigation
of whether differences in PCR error are detected between high fidelity
polymerases and number of cycles used. Second, for examining whether
these differences in error are enough to obscure smaller biological
signals than what has been previously reported (Gohl et al., 2016).
Thus, this study will investigate whether differences in high fidelity
DNA polymerases are due to error and whether these differences obscure
the biological variation in the fecal bacterial community composition
between individuals. We accomplish this by examining GC-based
amplification error, sequence error (e.g.~subsitutions), chimera
prevalence, the number of OTUs, Bray-Curtis index, and OTU-based
differences at varying number of cycles in five different high fidelity
DNA polymerases. Additionally, we also created Random Forest models to
classify number of cycles, polymerase, and individuals to assess which
model had the best classification success when using 16S rRNA gene
sequencing data.
